Sampling requirements to detect ecosystem change

### 2020-06-30 R/manuscript_draft.Rmd
Manuscript edits

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = dirname(getwd()))
# Optional code below to include package citations
 knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
# Use [@R-dplyr; @R-rmarkdown] in main text
```

@@

```{r load_packages,echo=F,warning=F,message=F}
if (!require("pacman",character.only = TRUE))
  {
    install.packages("pacman",dep=TRUE)
    if(!require("pacman",character.only = TRUE)) stop("Package not found")
  }

@@

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

@@

For both our simulations and case study, we investigate the effect of different sampling strategies on our ability to detect a changepoint. We begin by either creating simulated time series or using an actual paleoecological time series (Fig. \ref{fig:conceptual}). We then subsampled each time series [@White2020] to test the effect of three different sampling approaches along with varying the sample size (Fig. \ref{fig:conceptual}c-f). We compared the estimated changepoint from the subsampled time series to that of the full time series as a measure of the effectiveness.

@@

We examined the performance of our approach to detect changepoints in a paleosequence. Paleolimnology provides a reconstruction of past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances). A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).

We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage. Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed. Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018]. Air temperature is now driving changes in plankton communities [@Bruel2018].

In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using less samples. In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

@@

We found that random sampling performed the worst, as changepoint analysis was left to chance (Fig. \ref{fig:case_study_number_sample}). Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance). Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig. \ref{fig:case_study_number_sample}c). We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig. \ref{fig:case_study_iterative_sampling})

We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018). Iterative sampling still performed best, especially if an higher number of initial samples was chosen (Fig. S3).

As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). <!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication

>>>>

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = dirname(getwd()))
# Optional code below to include package citations
 knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
# Use [@R-dplyr; @R-rmarkdown] in main text

# Load packages
if (!require("pacman",character.only = TRUE))
  {
    install.packages("pacman",dep=TRUE)
    if(!require("pacman",character.only = TRUE)) stop("Package not found")
  }

@@

@@

Monitoring programs characteristics must be tightly linked to the questions of interest. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

@@

For both our simulations and case study, we investigate the effect of different sampling strategies on our ability to detect a changepoint. We begin by either creating simulated time series or using an actual paleoecological time series (Fig. \ref{fig:conceptual}). We then subsampled each time series to test the effect of three different sampling approaches along with varying the sample size (Fig. \ref{fig:conceptual}c-f). We compared the estimated changepoint from the subsampled time series to that of the full time series as a measure of the effectiveness.

@@

We examined the performance of our approach to detect changepoints in a paleosequence. Paleolimnology allows to reconstruct past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances). A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).

We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in zooplankton Cladocera assemblage. Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed. Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018]. Air temperature is now driving changes in plankton communities [@Bruel2018].

In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using less samples. In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

@@

We found that random sampling performed the worst, as changepoint analysis was left to chance (Fig. \ref{fig:case_study_number_sample}). Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., also left to chance). Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig. \ref{fig:case_study_number_sample}c). We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig. \ref{fig:case_study_iterative_sampling})

We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018). Iterative sampling still performed best, especially if an higher number of initial samples (7) was chosen (Fig. S3).

As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset. <!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication
